---
title: "Biophysical Prediction of HNF4A Binding Sites"
output: html_notebook
---

```{r}
library(DNAshapeR)
library(biomaRt)
library(dplyr)
library(tidyverse)
library(seqinr)
library(Biostrings)
library(TFBSTools)
library(seqLogo)
library(JASPAR2018)
library(reshape2)
library(ggplot2)
library(nucleR)
library(rtracklayer)
library(bedr)
library(coin)
library(car)
library(mltools)
library(caret)
library(InformationValue)
library(plotROC)
```

We need to first load up the DNA accesibility data and the HNF4A binding data from bed files.
To use this code you will need to have installed bedtools

```{r}
if(check.binary("bedtools")){
  dnaseq <- read.table(gzfile('../data/liver_euchormatin/ENCFF955WVT_HEPG2_DNaseSeq.bed.gz'), header = FALSE, stringsAsFactors = FALSE)
  hnf4a <- read.table(gzfile('../data/NR_NarrowPeaks/HN4A_peaks.bed'), header  = FALSE, stringsAsFactors = FALSE)
  
  #Check Files viability
  is.a.valid <- is.valid.region(dnaseq)
  is.b.valid <- is.valid.region(hnf4a)
  
  a <- dnaseq[is.a.valid,]
  b <- hnf4a[is.b.valid,]
  
  #Sort Files for bed tools speedup
  a.sort <- bedr.sort.region(a)
  b.sort <- bedr.sort.region(b)
  
  #Remove Redundant Regions
  a.merge <- bedr.merge.region(a.sort)
  b.merge <- bedr.merge.region(b.sort)

  #Doing Bed Arithmetic
  #For this I will intersect hnf4a and the hepg2 availibility data and then subtract the hnf4a intersection (positives) from the hepg2 data to get the complement (negatives)
  pos <- bedr.join.region(b.merge, a.merge)
  neg <- bedr.subtract.region(a.merge, pos)
  
  #Removing non-overlapping rows in intersection
  pos <- pos[!pos['start.b'] == -1, 1:4]
}
```
Now that we have our data, we will now load the sequences.

```{r}
#Load Genome
genome <- TwoBitFile('../data/hg38.2bit')

#Load Locations
pos_locs <- GRanges(pos$chr, IRanges(pos$start, pos$end))
neg_locs <- GRanges(neg$chr, IRanges(neg$start, neg$end))

#Add Sequence
pos$seq <- as.character(import(genome, which = pos_locs))
neg$seq <- as.character(import(genome, which = neg_locs))

#Bin Sequences
pos_bins <- c()
for (seq in pos$seq){
  string <- seq
  num.char <- nchar(string) 
  extra <- num.char%%100
  start <- ceiling(extra/2)
  if (num.char > 100){
    starts <- seq(start,num.char-100, by=100)
    pos_bins <- c(pos_bins, sapply(starts, function(i){
      substr(string, i+1, i+100)
      }
      )
      )
  }
}

neg_bins <- c()
for (seq in neg$seq){
  string <- seq
  num.char <- nchar(string) 
  extra <- num.char%%100
  start <- ceiling(extra/2)
  if (num.char > 100){
    starts <- seq(start,num.char-100, by=100)
    neg_bins <- c(neg_bins, sapply(starts, function(i){
      substr(string, i+1, i+100)
      }
      )
      )
  }
}
```


We will start with the Average PWM feature

```{r}
tf <-  	"MA0114.2"
pfm <-getMatrixByID(x =JASPAR2018, ID=tf)
pwm <- toPWM(pfm)
pwmScores <- c()
for (seq in  pos_bins){
  DNAStr <- DNAString(c2s(seq)) 
  pwmSearch <- searchSeq(pwm, DNAStr, min.score = 0)
  scores <- pwmSearch
  pwmScore <- mean(scores)
  pwmScores <- c(pwmScores, pwmScore)
}

positives <- data.frame(Sequence = pos_bins, avePWMScore = pwmScores)

pwmScores <- c()
for (seq in  neg_bins){
  DNAStr <- DNAString(c2s(seq)) 
  pwmSearch <- searchSeq(pwm, DNAStr, min.score = 0)
  scores <- relScore(pwmSearch)
  pwmScore <- mean(scores)
  pwmScores <- c(pwmScores, pwmScore)
}

negatives <- data.frame(Sequence = neg_bins, avePWMScore = pwmScores)
```

Now we will build the DNAShape Model

First We need to write our Sequences to a fasta file

```{r}
# Write Postive Sequences
names <- c()
for (i in 1:length(positives$Sequence)){
 names <- c(names, paste0("HNF4A_POS_", as.character(i)))
}

write.fasta(as.list(as.character(positives$Sequence)), names, file.out = "Documents/sudin_pipeline/data/hnf4aPos.fasta")

#Write Negative Sequences
names <- c()
for (i in 1:length(negatives$Sequence)){
 names <- c(names, paste0("HNF4A_NEG_", as.character(i))) 
}

write.fasta(as.list(as.character(negatives$Sequence)), names, file.out = "Documents/sudin_pipeline/data/hnf4aNeg.fasta")
```

Now We Want to Encode the DNA as Shape Features
```{r}
#Getting DNAShapeFeatures for Postive Sequences
posShape <- getShape("../data/hnf4aPos.fasta")
#Getting DNA
negShape <- getShape("../data/hnf4aNeg.fasta")
```
Getting the sequence shape vectors

```{r}
features <- c( "1-MGW","1-ProT", "1-Roll", "1-HelT")
posShapeVec <- encodeSeqShape("../data/hnf4aPos.fasta", posShape, featureNames = features, normalize = TRUE)
negShapeVec <- encodeSeqShape("../data/hnf4aNeg.fasta", posShape, featureNames = features, normalize = TRUE)
```

Getting the means for all these features

```{r}
positives$MGW <- rowMeans(posShape$MGW, na.rm = TRUE)
positives$HelT <- rowMeans(posShape$HelT, na.rm = TRUE)
positives$ProT <- rowMeans(posShape$ProT, na.rm = TRUE)
positives$Roll <- rowMeans(posShape$Roll, na.rm = TRUE)
positives$EP <- rowMeans(posShape$EP, na.rm = TRUE)

negatives$MGW <- rowMeans(negShape$MGW, na.rm = TRUE)
negatives$HelT <- rowMeans(negShape$HelT, na.rm = TRUE)
negatives$ProT <- rowMeans(negShape$ProT, na.rm = TRUE)
negatives$Roll <- rowMeans(negShape$Roll, na.rm = TRUE)
negatives$EP <- rowMeans(negShape$EP, na.rm = TRUE) 
```

Previewing Signifigance Distributions Using Approximate Fisher-Pittman
```{r}
positives$y <- as.factor(1)
negatives$y <- as.factor(0)

concat <- rbind(positives, negatives)
res1 <- oneway_test(HelT~y,concat[,2:8], distribution="approximate")
res2 <- oneway_test(ProT~y,concat[,2:8], distribution="approximate")
res3 <- oneway_test(MGW~y,concat[,2:8], distribution="approximate")
res4 <- oneway_test(EP~y,concat[,2:8], distribution="approximate")
res5 <- oneway_test(Roll~y,concat[,2:8], distribution="approximate")
res6 <- oneway_test(avePWMScore~y,concat[,2:8], distribution="approximate")
```

Figuring Out Full model
```{r}
positives <- data.frame(posShapeVec)
negatives <- data.frame(negShapeVec)

positives$y <- 1
negatives$y <- 0

concat <- rbind(positives, negatives)
```


Benchmarking with Logistic Regression
```{r}
#Getting Training and Testing Sets
set.seed(42)
smp_size <-  floor(0.75*nrow(concat))
train_ind <- sample(seq_len(nrow(concat)), size = smp_size)

train <- concat[train_ind, ]
test <- concat[-train_ind, ]
```


```{r}
model <- glm(y~., data = train, family = binomial(link = logit))
predicted <- plogis(predict(model, test))
optCutOff <- optimalCutoff(test$y, predicted)[1] 
```

```{r}
summary(model)
```

Check for multicolinearity
```{r}
vif(model)
```

This is worrisome for the model. Variance Inflation factors (vif) are well above 4 when I include all of the data.

When I take away MGW and ProT, I bring the vif ~ 10, which is somewhat acceptable.

MCC, confusion matrix, roc_curve
```{r}
confusionMatrix(test$y, predicted, threshold = optCutOff)
```


```{r}
plotROC(test$y, predicted)
```

Try with RFC?

```{r}

```

