---
title: "Computational Search for Toxicologically Relevant NR's TF's using Machine Learning"
output: html_notebook
---

The genes of interest are:
CYP … 3A4, 2C9, 2C8, 2E1, 1A2

2A6, 2D6, 2B6, 2C19, 3A5

2J2, 1A1, and 1B1

Agnostic
Pipeline:
Download Encode Acessible Regions in Liver ->
Extract Sequences using BED and DEEP TOOL -> 
DREME-Chip Search for Binding Motifs ->
Cross Reference With ChipSeq Data -> 
Visualize using cytoscape

Literature Based
Pipeline:
Download Accessible Regions Using Cytoscape -> 
Extract Sequencess using BED and DEEP TOOL -> 
Cross Reference with JASPAR with Known TF's ->
Cytoscape visualizaiton

This will be the proof of concept
Eventually we want to replace Step 3 of each pipeline with a Machine Learning Classifier

Possible Future Ideas:
TF Prediction -> SNP cataloging -> Precision Medicine Experimental Validation

SNP Predictive Algorithms?


Extracting Transcription Binding Candidates (TBC)

Setting Up dependencies
```{r}
library(biomaRt)
library(dplyr)
library(tidyverse)
library(seqinr)
library(Biostrings)
library(TFBSTools)
library(seqLogo)
library(JASPAR2018)
library(reshape2)
library(ggplot2)
library(nucleR)
library(rtracklayer)
```
Getting Gene Coordinates
```{r}
mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", host="grch37.ensembl.org", path="/biomart/martservice", dataset="hsapiens_gene_ensembl")
#Genes of interest
cyp <- c('ENSG00000160868','ENSG00000138109','ENSG00000138115','ENSG00000130649','ENSG00000140505','ENSG00000255974','ENSG00000100197','ENSG00000197408','ENSG00000165841','ENSG00000106258','ENSG00000134716','ENSG00000140465','ENSG00000138061')
#Getting gene annotation
ann <- getBM(c("hgnc_symbol","description","chromosome_name","band","strand","start_position","end_position","ensembl_gene_id"), "ensembl_gene_id", cyp, mart)
```

```{r}
gen_loc <- ann  %>%
  select(c('chromosome_name', 'start_position', 'end_position', 'ensembl_gene_id'))
gen_loc$chromosome_name <- paste('chr', gen_loc$chromosome_name, sep = '')
```

Loading DNA accessibility data from following Encode Data Sets:
ENCSR149XIL
ENCSR555QAY
ENCSR000ENP
ENCSR000DCH
ENCSR373TDL
ENCSR000DCW

```{r}
file <- '~/Documents/sudin_pipeline/data/liver_euchormatin/ENCFF435SDA_Hg19_HepG2_DNaseISeq.bed.gz'

nar_pek_var <- c('chrom', 'chromstart', 'chromend', 'name', 'score','strand', 'signalvalue', 'pvalue', 'qvalue', 'peak')
tbc <- read_tsv(file, col_names = nar_pek_var)
```
All at once relevant TBC for each gene
```{r}
cyp_reg <- c()
for (i in 1:length(gen_loc$chromosome_name)){
  j <- filter(tbc, chrom == gen_loc[[i, 1]] & chromstart > gen_loc[[i, 2]]-10000 & chromend < gen_loc[[i, 3]]+10000) 
  cyp_reg <- c(cyp_reg, j)
}
```
Building frequency tibble for plotting to find appropriate proximal upstream and downstream lengths for calculating
```{r}
bounds <- seq(1000, 20000, 1000)
dat <- tibble()
for (i in bounds){
  for (j in 1:length(gen_loc$chromosome_name)){
    r <- nrow(filter(tbc, chrom == gen_loc[[j, 1]] & chromstart > gen_loc[[j, 2]]-i & chromend < gen_loc[[j, 3]]+i))
    dat <- bind_rows(dat, c(freq = r, bound = i, gene = gen_loc[[j, 4]]))
  }
}
dat$bound <- as.numeric(dat$bound)
dat$freq <- as.numeric(dat$freq)
```
Building stacked bargraph to visualize TBC frequency at differing proximal lengths
```{r}
ggplot(dat, aes(x = bound,y = freq,  fill = gene)) + 
  geom_histogram(stat = 'identity') +
  ylab("DHS Relative Frequency") +
  xlab("Downstream/Upstream Region Size (bps)")
```

```{r}
ggplot(dat, aes(x = bound,y = freq/50,  fill = gene)) + 
  geom_histogram(stat = 'identity') +
  ylab("DHS Relative Frequency") +
  xlab("Downstream/Upstream Region Size (bps)")
```
I will use 10K bps as my proximal regions of interest.

So  now I want to use Jaspar to cross refercene TF's of interest against the regions of interest.
I now need to write up my TF's of interest and then use Jaspar to find those motifs.

In the meantime we can use Meme-Chip to find these Denovo

So now we need to create a system that will download the DNA based on some gene coordinates [chr[i], start, end]

To do that we will read from 2bit files to reduce memory strain

Test getting sequence from genome 
```{r}
genome <- TwoBitFile('../data/hg19.2bit')
test_loc <- GRanges('chr10', IRanges(12129441, 12129591))
test_seq <- as.character(import(genome, which = test_loc))
```
Trying to get all regulatory regions
```{r}
reg_seqs_vec <- tibble()
for (i in 1:length(gen_loc$chromosome_name)){
  j <- filter(tbc, chrom == gen_loc[[i, 1]] & chromstart > gen_loc[[i, 2]]-10000 & chromend < gen_loc[[i, 3]]+10000) 
  gene <- gen_loc[[i,4]]
  locs <- GRanges(j$chrom, IRanges(j$chromstart, j$chromend))
  seqs <- as.character(import(genome, which = locs))
  if (length(seqs) > 0){
    reg_seqs_vec <- bind_rows(reg_seqs_vec, tibble(rep(gene,length(seqs)), seqs))
  }
}
colnames(reg_seqs_vec) <- c('ensemblID', 'Sequence')
```
Getting Distal Elements
```{r}
tbde <- read_tsv('../data/allGeneCorrelations100000.p05_v3.txt')
cyp_names <- c('CYP3A4','CYP2C9','CYP2C8','CYP2E1','CYP1A2','CYP2A6','CYP2D6','CYP2B6','CYP2C19','CYP3A5','CYP2J2','CYP1A1','CYP1B1')
de <- tbde %>%
  filter(grepl(pattern = paste(cyp_names,collapse ='|'), x = gene_name, ignore.case = TRUE)) %>%
  filter(!gene_name == 'CYP3A43') %>%
  filter(pval < 0.05)
```
```{r}
ensemblID <- tibble(de$ensemblID)
loc <- GRanges(de$dhs_chr, IRanges(de$dhs_start,de$dhs_end))
sequences <- tibble(as.character(import(genome, which = loc)))
de_seqs_vec <- cbind(ensemblID, sequences)
colnames(de_seqs_vec) <- c('ensemblID', 'sequence')
```

Finding Motifs from Jaspar for relevant TFS:
ESR1, NR3C1, RARA, HNF4A, CAR, PXR, VDR, PPARA, AHR
CAR abd PXR were not found from JASPAR 2018 : Will need to see

I want to get the data for the TF's above

I want to get all of them
This script is missing CAR, PXR
```{r}
#Getting PWM for each TF binding site
tf <- c('ESR1', 'NR3C1', 'RARA','RARA(var.2)', 'RARA::RXRA', 'PPARA::RXRA', 'HNF4A', 'VDR','RXRA::VDR', 'Ahr::Arnt')
tf_pfm <- c()
tf_pwm <- c()
for (i in 1:length(tf)){
  tf_pfm <- getMatrixByName(JASPAR2018, tf[i])
  tf_pwm <- c(tf_pwm, toPWM(tf_pfm))
}
#For CAR NR
#Hosada et. al
require(ggseqlogo)
hosadaFile <- '../data/car_binding_sites_hosoda.tsv'
hosadaFileHeader <- c('number', 'loc', 'len', 'bs', 'motif', 'compete10', 'compete100')
hCarHosada <- read_tsv(hosadaFile, col_names = hosadaFileHeader, skip = 1)
hCarHosada <- filter(hCarHosada,  !is.na(bs))
hCarHosada <- mutate_all(hCarHosada, .funs=toupper)
siteSeqs <- hCarHosada$bs
#Make Postion Frequency Matrix
countMatrix <- consensusMatrix(siteSeqs)
pfm <- PFMatrix(ID = 'CAR', name = 'CAR-RXRA', profileMatrix = countMatrix)
ggseqlogo(countMatrix)
pwm <- toPWM(pfm)
tf_PWM <- c(tf_pwm, pwm)
```
Putting all the data together for analysis
```{r}
#Concatenating all my data
#DE Data
cyp_reg_data <- tibble(reg_gene = de_seqs_vec$ensemblID,
                       seq = de_seqs_vec$sequence,
                       dhs_chrom = de$dhs_chr,
                       dhs_start = de$dhs_start,
                       dhs_end = de$dhs_end,
                       de = 1
)
#+/- 10000 bp region
for (i in 1:length(gen_loc$chromosome_name)){
  j <- filter(tbc, chrom == gen_loc[[i, 1]] & chromstart > gen_loc[[i, 2]]-10000 & chromend < gen_loc[[i, 3]]+10000) 
  gene <- gen_loc[[i,4]]
  locs <- GRanges(j$chrom, IRanges(j$chromstart, j$chromend))
  seqs <- as.character(import(genome, which = locs))
  if (length(seqs) > 0){
    k <- tibble(reg_gene = gen_loc$ensembl_gene_id[i],
           seq = seqs,
           dhs_chrom = j$chrom,
           dhs_start = j$chromstart,
           dhs_end = j$chromend,
           de = 0
    )
    cyp_reg_data <- rbind(cyp_reg_data, k)
  }
}
```

Since we are illustrating the problem we will use unaltered PWMs
There will be a relaxed PWM to create a wider "net"
```{r}
tf <- c('ESR1', 'NR3C1', 'RARA','RARA(var.2)', 'RARA::RXRA', 'PPARA::RXRA', 'HNF4A', 'VDR','RXRA::VDR', 'Ahr::Arnt')
frequency <- c()
for (p in 60:90){
  longSites <- tibble(seq = character(),
                    chrom = character(),
                    start_coord = integer(),
                    end_coord = integer(),
                    ensembl = character(),
                    de = integer(),
                    tf = character()
                    )
  for (i in 1:length(tf)){
    #Getting PWM from Jaspar2018
    tf_pfm <- getMatrixByName(JASPAR2018, tf[i])
    tf_pwm <- toPWM(tf_pfm)
    TF <- tf[i]
    for (j in 1:length(cyp_reg_data$seq)){
      #Define Gene of interest
      gene <- cyp_reg_data$reg_gene[j]
      #Defining Peak Location
      peak_chrom2 <- cyp_reg_data$dhs_chrom[j]
      peak_start <- cyp_reg_data$dhs_start[j]
      # Search sequence for relaxed match to each tf    
      mySeqDNAString <- DNAString(c2s(cyp_reg_data$seq[j]))
      siteSet <- searchSeq(tf_pwm, mySeqDNAString, 
                           seqname = "seq1", min.score = paste0(as.character(p),"%"), strand = "*")
      sites_in_myseq <- writeGFF3(siteSet)
      de_sing <-  cyp_reg_data$de[j]
      # If binding sites found, add all instances to dataframe of sites
      numSites_in_myseq <- dim(sites_in_myseq)[1]
      if (numSites_in_myseq > 0) {
        cat("* Found ", numSites_in_myseq, " putative REs in peak # ", j, " \n")
        
        # Loop over DREs
        for (k in 1:numSites_in_myseq) {
          # Find start and end coords of site
          start <- sites_in_myseq$start[k]
          end <- sites_in_myseq$end[k]
          
          # Ignore sites too close to end of sequence string
          if ((start <= 7) || (end >= length(mySeqDNAString) - 7)) {
            next
          }
          
          if (sites_in_myseq$strand[k] == "+") {
            long_site <- mySeqDNAString[(start - 7):(end + 7)]
          } else {
            long_site <- reverseComplement(mySeqDNAString[(start - 7):(end + 7)])
          }
          #cat("DRE no. = ", j, "\n")
          long_site_string <- toString(long_site)
          start_long_site = (start - 7)
          end_long_site = (end + 7)
          temp_df <- data.frame(seq = long_site_string, 
                                chrom = peak_chrom2,
                                start_coord = start_long_site + peak_start, 
                                end_coord = end_long_site + peak_start,
                                ensembl = gene,
                                de = de_sing,
                                tf = TF
                                )
          longSites <- rbind(longSites, temp_df) 
        }
      }
    }
  }
  frequency <- c(frequency,length(longSites$seq))
}
```
Creating a Frequency Plot
```{r}
df <- tibble(
  thresh = 60:90,
  freq = frequency
)
ggplot(df,aes(x = thresh, y = freq)) +
  geom_bar(fill = "#0073C2FF", stat = "identity")
```
Based on the Cutoff Analysis that I have performed, I suggest a match score of 82%
```{r}
tf <- c('Ahr::Arnt', 'ESR1', 'HNF4A', 'NR3C1', 'PPARA::RXRA', 'RARA', 'VDR')
frequency <- c()

longSites <- tibble(seq = character(),
                  chrom = character(),
                  start_coord = integer(),
                  end_coord = integer(),
                  ensembl = character(),
                  de = integer(),
                  tf = character()
                  )
for (i in 1:length(tf)){
  #Getting PWM from Jaspar2018
  tf_pfm <- getMatrixByName(JASPAR2018, tf[i])
  tf_pwm <- toPWM(tf_pfm)
  TF <- tf[i]
  for (j in 1:length(cyp_reg_data$seq)){
    #Define Gene of interest
    gene <- cyp_reg_data$reg_gene[j]
    #Defining Peak Location
    peak_chrom2 <- cyp_reg_data$dhs_chrom[j]
    peak_start <- cyp_reg_data$dhs_start[j]
    # Search sequence for relaxed match to each tf    
    mySeqDNAString <- DNAString(c2s(cyp_reg_data$seq[j]))
    siteSet <- searchSeq(tf_pwm, mySeqDNAString, 
                         seqname = "seq1", min.score = "80%", strand = "*")
    sites_in_myseq <- writeGFF3(siteSet)
    de_sing <-  cyp_reg_data$de[j]
    # If binding sites found, add all instances to dataframe of sites
    numSites_in_myseq <- dim(sites_in_myseq)[1]
    if (numSites_in_myseq > 0) {
      cat("* Found ", numSites_in_myseq, " putative REs in peak # ", j, " \n")
      
      # Loop over DREs
      for (k in 1:numSites_in_myseq) {
        # Find start and end coords of site
        start <- sites_in_myseq$start[k]
        end <- sites_in_myseq$end[k]
        
        # Ignore sites too close to end of sequence string
        if ((start <= 7) || (end >= length(mySeqDNAString) - 7)) {
          next
        }
        
        if (sites_in_myseq$strand[k] == "+") {
          long_site <- mySeqDNAString[(start - 7):(end + 7)]
        } else {
          long_site <- reverseComplement(mySeqDNAString[(start - 7):(end + 7)])
        }
        #cat("DRE no. = ", j, "\n")
        long_site_string <- toString(long_site)
        start_long_site = (start - 7)
        end_long_site = (end + 7)
        temp_df <- data.frame(seq = long_site_string, 
                              chrom = peak_chrom2,
                              start_coord = start_long_site + peak_start, 
                              end_coord = end_long_site + peak_start,
                              ensembl = gene,
                              de = de_sing,
                              tf = TF
                              )
        longSites <- rbind(longSites, temp_df) 
      }
    }
  }
}
```
 

Doing CAR Binding Site Search
```{r}
#Data used to check how well my CAR pwm is up to snuff
mm10 <- TwoBitFile('../data/mm10.2bit')
# Chip Exo Data Positions
hCarChipBed <- '../data/NR_NarrowPeaks/CAR_PB_peaks.bed.gz'
hCarChipBedColNames <- c('chrom', 'chromstart', 'chromend', 'name_site', 'score')
hCarChipLoc <- read_tsv(hCarChipBed, col_names = hCarChipBedColNames)
#Extracting sequences
j <- head(hCarChipLoc)
hCarChipSeqs <- c()
locs <- GRanges(j$chrom, IRanges(j$chromstart, j$chromend))
seqs <- as.character(import(mm10, which = locs))
rm(j)
hCarChipLoc$Seqs <- seqs
rm(seqs)

```

```{r}
CARSites <- tibble(seq = character(),
                   peak = integer(),
                    chrom = character(),
                    start_coord = integer(),
                    end_coord = integer(),
                    tf = character()
                    )
for (i in 1:length(hCarChipLoc$chrom)){
    #Defining Peak Location
    peak_chrom2 <- hCarChipLoc$chrom[i]
    peak_start <- hCarChipLoc$chromstart[i]
    # Search sequence for relaxed match to each tf    
    mySeqDNAString <- DNAString(c2s(hCarChipLoc$Seqs[i]))
    siteSet <- searchSeq(pwm, mySeqDNAString, 
                         seqname = "seq1", min.score = "70%", strand = "*")
    sites_in_myseq <- writeGFF3(siteSet)
    # If binding sites found, add all instances to dataframe of sites
    numSites_in_myseq <- dim(sites_in_myseq)[1]
    if (numSites_in_myseq > 0) {
      # Loop over DREs
      for (j in 1:numSites_in_myseq) {
        # Find start and end coords of site
        start <- sites_in_myseq$start[j]
        end <- sites_in_myseq$end[j]
        
        # Ignore sites too close to end of sequence string
        if ((start <= 7) || (end >= length(mySeqDNAString) - 7)) {
          next
        }
        
        if (sites_in_myseq$strand[j] == "+") {
          long_site <- mySeqDNAString[(start - 7):(end + 7)]
        } else {
          long_site <- reverseComplement(mySeqDNAString[(start - 7):(end + 7)])
        }
        #cat("DRE no. = ", j, "\n")
        long_site_string <- toString(long_site)
        start_long_site = (start - 7)
        end_long_site = (end + 7) 
        temp_df <- data.frame(seq = long_site_string,
                              peak = i,
                              chrom = peak_chrom2,
                              start_coord = start_long_site + peak_start, 
                              end_coord = end_long_site + peak_start,
                              tf = 'CAR:RXRA'
                              )
        CARSites <- rbind(CARSites, temp_df) 
      }
    }
  }
```

So I got the data but I still need to test it. It seems pretty sensitive so I'm going to do an ROC curve on the data to see how well it performs on that specific data. I acutally should do this for all of my data to see how well all of them perform. This is to get a base line of the performance of these PWM's on these datasets. 

This should be our standard of comparison for our machine learning models

```{r}
#Retrieving Benchmark data
#Made of 200 Peaks with scores above 100, randomly selected
#Exception for NR3C1 which has only 55 peaks, some of them low quality
bench_dat_file <- '../data/NR_NarrowPeaks/testSetPwm_2.bed'
bench_header <- c(nar_pek_var, 'TF')
bench_dat <- read_tsv(bench_dat_file, col_names = bench_header)
hg19locs <- bench_dat %>%
  filter(TF == 'PPARA')
mm10locs <- bench_dat %>%
  filter(TF == 'CAR')
hg38locs <- bench_dat %>%
  filter(TF != 'CAR' & TF != 'PPARA')
#Data analysis was performed on GChr38 so I will use the hg38.2bit file as my ref genome
hg19 <- TwoBitFile('../data/hg19.2bit')
mm10 <- TwoBitFile('../data/mm10.2bit')
hg38 <- TwoBitFile('../data/hg38.2bit')
#Will now retrieve sequences from reference. 
hg19_locs <- GRanges(hg19locs$chrom, IRanges(hg19locs$chromstart, hg19locs$chromend))
mm10_locs <- GRanges(mm10locs$chrom, IRanges(mm10locs$chromstart, mm10locs$chromend))
hg38_locs <- GRanges(hg38locs$chrom, IRanges(hg38locs$chromstart, hg38locs$chromend))
hg19locs$seqs <- as.character(import(hg19, which = hg19_locs))
mm10locs$seqs <- as.character(import(mm10, which= mm10_locs))
hg38locs$seqs <- as.character(import(hg38, which = hg38_locs))
bench_dat <-  rbind(hg38locs, mm10locs, hg19locs)
rm(hg38locs, hg19locs, mm10locs, hg19_locs, hg38_locs, mm10_locs)
```

```{r}
#Creating data for ROCAUC plots
tf <- c('ESR1', 'NR3C1', 'RARA','RARA(var.2)', 'RARA::RXRA', 'PPARA::RXRA', 'HNF4A', 'VDR','RXRA::VDR', 'Ahr::Arnt', 'CAR')
tf_re <- c('ESR1', 'NR3C1', 'RARA', 'RARA', 'RARA:RXRA', 'PPARA::RXRA', 'HN4A', 'D3', 'RXRAA:D3', 'AHR', 'CAR')
for (i in 1:length(tf)){
  if tf != 
  mutate(bench_dat)
}
#Example of how to create binary data
mutate(bench_dat, true.val = ifelse(TF == 'ESR1', 1, 0))$true_val
#Come back to this, it may come up later for benchmarking
```
I need to export some TFBS as PWM for my code
```{r}
nr2 <- 'Nuclear Receptor class'
tf_pfm <- getMatrixByName(JASPAR2018, nr2)
tf_pwm <- toPWM(tf_pfm)
```

